Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations

### 2021-11-23 analysis/paper/paper.Rmd
added Nick's remaining plots to ms

One potentially important, but perhaps under-appreciated source of mortality for Chinook Salmon is predation on emigrating juveniles by piscivorous fishes, including both native and non-native species. Dams and reservoirs in the Columbia River basin are the primary locations associated with high rates of predation on salmonids [@Petersen1994; @Ward1995]. There are generally two maechanisms that explain these high predation zones. First, migration rates of juvenile Chinook Salmon are reduced during reservoir passage [@Venditti2000], thereby increasing the time migrating smolts are vulnerable to predation. Second, reservoirs and downstream tailraces created by dams on the Columbia and Snake rivers have created favorable slow-water habitat for several species of piscivorous fish known to consume juvenile Chinook Salmon including Walleye *Sander vitreus*, Smallmouth Bass *Micropterus dolomieu*, and Northern Pikeminnow *Ptychocheilus oregonensis*. In the Lower Granite Reservoir on the Snake River, juvenile Chinook Salmon are common prey items of non-native Smallmouth Bass during some seasons [@Erhardt2018]. Walleye also prefer slow water habitats and are considered a potentially important source for Juvenile Chinook Salmon mortality [@NOAA2020], particularly as their abundance has significantly increased in Snake River dam facilities between 2008 and 2020.

In addition to non-native species, Northern Pikeminnow are a native, piscivorous fish preferring slow water habitats. Consequently, they too have benefited from dams on the Columbia River and become abundance predators on salmonid outmigrants [@Knutsen1999]. Indeed, Northern Pikeminnow are estimated to consume 8% (16.4 million) of the estimated 200 million juvenile salmonids emigrating through the lower Snake and Columbia rivers, annually [@Beamesderfer1996]. Predation by piscivorous fish may have been a factor in the below-average survival of wild Chinook Salmon in 2020 [@NOAA2020], however there is a dearth of current data on piscivorous predator populations in the Columbia River.

@@

```{r gastric-data}
gastric_df = raw_df %>%
  filter(species == "Northern Pikeminnow") %>%
  mutate(fish_or_parts = case_when(stri_detect_fixed(stomach_contents, "Fish") ~ "Fish or Fish Parts",
                                   stri_detect_fixed(stomach_contents, "Empty") ~"Empty",
                                   stri_detect_fixed(comments, "Whole scuplin") ~ "Fish or Fish Parts",
                                   stri_detect_fixed(stomach_contents, "Whitefish") ~ "Fish or Fish Parts",
                                   TRUE ~ "Other")) %>%
  drop_na(length) %>%
  mutate(non_fish_wt = stomach_contents_weight - fish_content_weight) %>%
  dplyr::select(date, length, lavage_id, stomach_contents, comments, fish_or_parts, stomach_contents_weight, non_fish_wt,                  fish_content_weight) %>%
  filter(!stomach_contents %in% c("Lost sample CG27-001", "Did Not Lavage")) %>%
  drop_na(stomach_contents)

@@

>>>>

One potentially important, but perhaps under-appreciated source of mortality for Chinook Salmon is predation on emigrating juveniles by piscivorous fishes, including both native and non-native species. Dams and reservoirs in the Columbia River basin are the primary locations associated with high rates of predation on salmonids [@Petersen1994; @Ward1995]. There are generally two maechanisms that explain these high predation zones. First, migration rates of juvenile Chinook Salmon are reduced during reservoir passage [@Venditti2000], thereby increasing the time migrating smolts are vulnerable to predation. Second, reservoirs and downstream tailraces created by dams on the Columbia and Snake rivers have created favorable slow-water habitat for several species of piscivorous fish known to consume juvenile Chinook Salmon including Walleye *Sander vitreus*, Smallmouth Bass *Micropterus dolomieu*, and Northern Pikeminnow *Ptychocheilus oregonensis*. In the Lower Granite Reservoir on the Snake River, juvenile Chinook Salmon are common prey items of non-native Smallmouth Bass during some seasons [@Erhardt2018]. Walleye also prefer slow water habitats and are considered a potentially important source for Juvenile Chinook Salmon mortality [@Widener2021], particularly as their abundance has significantly increased in Snake River dam facilities between 2008 and 2020.

In addition to non-native species, Northern Pikeminnow are a native, piscivorous fish preferring slow water habitats. Consequently, they too have benefited from dams on the Columbia River and become abundance predators on salmonid outmigrants [@Knutsen1999]. Indeed, Northern Pikeminnow are estimated to consume 8% (16.4 million) of the estimated 200 million juvenile salmonids emigrating through the lower Snake and Columbia rivers, annually [@Beamesderfer1996]. Predation by piscivorous fish may have been a factor in the below-average survival of wild Chinook Salmon in 2020 [@Widener2021], however there is a dearth of current data on piscivorous predator populations in the Columbia River.

@@

```{r gastric-data}
gastric_df = raw_df %>%
  filter(species == "Northern Pikeminnow") %>%
  mutate(fish_or_parts = case_when(stri_detect_fixed(stomach_contents, "Fish") ~ "Fish or Fish Parts",
                                   stri_detect_fixed(stomach_contents, "Empty") ~"Empty",
                                   stri_detect_fixed(comments, "Whole scuplin") ~ "Fish or Fish Parts",
                                   stri_detect_fixed(stomach_contents, "Whitefish") ~ "Fish or Fish Parts",
                                   TRUE ~ "Other")) %>%
  drop_na(length) %>%
  mutate(non_fish_wt = stomach_contents_weight - fish_content_weight) %>%
  dplyr::select(date, method, length, lavage_id, stomach_contents, comments, fish_or_parts, stomach_contents_weight, non_fish_wt, fish_content_weight) %>%
  filter(!stomach_contents %in% c("Lost sample CG27-001", "Did Not Lavage")) %>%
  drop_na(stomach_contents)

# hungry_fish_df = gastric_df %>%
#   filter(fish_or_parts == "Fish or Fish Parts") %>%
#   mutate(fish = case_when(stri_detect_fixed(stomach_contents, "Shiner") ~ "Redside Shiner",
#                           stri_detect_fixed(stomach_contents, "Whitefish") ~ "Mountain Whitefish",
#                           stri_detect_fixed(stomach_contents, "scuplin") ~ "Sculpin",
#                           stri_detect_fixed(stomach_contents, "Chinook") ~ "Chinook",
#                           stri_detect_fixed(comments, "whitefish") ~ "Mountin Whitefish",
#                           stri_detect_fixed(comments, "scuplin") ~ "Sculpin",
#                           stri_detect_fixed(comments, "chinook") ~ "Chinook",
#                           TRUE ~ "Other"))

```{r gastric-summ}
gastric_df$npm_50 = lencat(gastric_df$length, w = 50)

psd_df = gastric_df %>%
  group_by(npm_50) %>%
  summarize(n = n())

gastric_df2 = gastric_df %>%
  drop_na(stomach_contents_weight) %>%
  group_by(fish_or_parts)

stmch_ctgry_summ = gastric_df2 %>%
  count()

stmch_wt_summ = gastric_df2 %>%
  summarise(sum(stomach_contents_weight))

non_fish_wt_summ = gastric_df2 %>%
  summarise(sum(non_fish_wt))

fish_wt_summ = gastric_df2 %>%
  summarise(sum(fish_content_weight))

@@

```{r stmch-cont-p, eval = F, fig.cap = "A stomach contents plot."}
stmch_cont_per_sz_p = gastric_df %>%
  filter(method %in% "Angling") %>%
  ggplot() +
  aes(x = fish_or_parts) +
  geom_bar(fill = "#DB740E") +
  labs(x = "Stomach Contents",
       y = "Frequency") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5,
                                   hjust = 1)) +
  facet_wrap(vars(npm_50),
             nrow = 2L)

stmch_cont_per_sz_p

```{r length-freq-p, eval = T, fig.cap = "A length frequency plot."}
length_freq_p = gastric_df %>%
  ggplot(aes(length)) +
  geom_histogram(breaks = seq(150, 650, by = 50),
                 col = "#DB740E",
                 fill = "#DB740E",
                 alpha = .2) +
  labs(x = "TL (mm)",
       y = "Frequency") +
  xlim(c(100, 700)) +
  ylim(c(0, 800)) +
  scale_x_continuous(breaks = seq(0, 650, by = 50)) +
  scale_y_continuous(breaks = seq(0, 550, by = 50)) +
  theme_classic()

length_freq_p
